From Reversible Programs to \\ Univalent Universes and Back

### 2021-10-30 Pi/paper/applications.tex
Copy-editing applications section

Using our semantics, we can normalise, synthesise, prove equivalence, and reason about~$\PiLang$ programs.

@@

Normalisation happens by evaluating a combinator to a permutation and then reifying it back. The evaluation step goes
from $\PiLang$ to $\PiPlusLang$ reducing $\times$ to $+$ and translating the corresponding combinators, then from
$\PiPlusLang$ to $\PiHatLang$ reducing them to natural numbers and adjacent transpositions, then finally to $\UFin$
through words in $\Sn$ and Lehmer codes. Then, the quotation step goes all the way back until $\PiPlusLang$. We define
this composite $\normt$ function below using our previous evaluation $\gdenot{\blank}$.

@@

As an example, consider the \Afun{reversibleOr1} circuit introduced in~\cref{sec:pi}. After applying $\normt_{1}$, the
type $\mathbb{B}\ 3$ normalises to $\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
  (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$, and the circuit reduces to the following
normal form:

@@

After applying $\normt_{1}$ to each circuit, we verify that they both reduce to the same normal form
\Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
these, we get the 2-combinator witnessing their equivalence.

@@

Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even permutations
always produces an even permutation, which answers the question to the negative.

Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
$c$. It is implemented first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$
defined in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the length of the list.

\begin{propositionrep}
  \label{prop:parity-preserved}
  For two $\PiLang$ circuits $c_1, c_2$, if there exists a 2-combinator between them, then $\parity c_1 = \term{parity} c_2$. 
\end{propositionrep}
\begin{proof}
  First, it is easy to check that Coxeter relations preserve parity of the list. Second, if $c_1, c_2$ are related by a
  2-combinator, then their list-of-transpositions representations are related in the Coxeter relation. Composing these
  two facts completes the proof.
\end{proof}

Using ~\cref{prop:parity-preserved} and showing that any composition of even cicuits yields an even ciruit, we can
answer the question by computing the parity of the \Afun{toffoli₄} and all of the \Afun{toffoli₃} gates.

Thus, we formulated the theorem using the representation on lists of transpositions, and transferred it to circuits using
the equivalences proven in the previous sections.

% More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
% programs.

>>>>

Using our semantics, we can normalise, synthesise, prove equivalence, and reason about~$\PiLang$ programs. We describe a
few applications in this section, and a more comprehensive list of applications and examples along with their
foramlisation is available in our artifact~\cite{choudhuryArtifactSymmetriesReversible2021}.

@@

Normalisation happens by evaluating a combinator to a permutation and then reifying it back. The evaluation step goes
from $\PiLang$ to $\PiPlusLang$ reducing $\times$ to $+$ and translating the corresponding combinators, then from
$\PiPlusLang$ to $\PiHatLang$ reducing them to natural numbers and adjacent transpositions, then finally to $\UFin$
through words in $\Sn$ and Lehmer codes. Finally, the quotation step goes all the way back until $\PiPlusLang$. We
define this composite $\normt$ function below using our previous evaluation function $\gdenot{\blank}$.

@@

\noindent
As an example, consider the \Afun{reversibleOr1} circuit introduced in~\cref{sec:pi}. After applying $\normt_{1}$, the
type $\mathbb{B}\ 3$ normalises to $\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
  (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$, and the circuit reduces to the following
normal form:

@@

After applying $\normt_{1}$ to each circuit, we verify that they both reduce to the same normal form
\Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
these, we get the 2-combinator witnessing their equivalence. More general, user-guided, reasoning can be done using the
sound and complete level-2 combinators to rewrite $\PiLang$ programs.

@@

Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃}
gates implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even
permutations always produces an even permutation, which answers the question to the negative. Thus, we want to formulate
the theorem using the representation on lists of transpositions, and transfer it to circuits using the equivalences
proven in the previous sections.

Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a
combinator $c$, by first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the
partiy of the length of the list.

\begin{propositionrep}
  \label{prop:parity-preserved}
  For two $\PiLang$ circuits $c_1, c_2$, if $c_1 \Iso c_2$, then $\parity(c_1) = \term{parity}(c_2)$.
\end{propositionrep}
\begin{proof}
  First, it is an easy check that Coxeter relations preserve parity of the list. Second, if $c_1$ and $c_2$ are related
  by a 2-combinator, then their Lehmer codes are the same, so their list-of-transpositions representations have to be
  related by the Coxeter relation. Composing these two facts completes the proof.
\end{proof}

Using ~\cref{prop:parity-preserved} and already mentioned lemma that any compostion of even circuits is even, we can
finish the proof by running $\parity$ on \Afun{toffoli₄} and on all of the \Afun{toffoli₃}.
